---
title: "Assignment 2"
author: "Anna Lai"
date: "March 07, 2023"
output:
  html_document:
    toc: yes
    df_print: paged
---

# Differential Gene Expression and Preliminary ORA

## Setup

Package installation commands hided. 
```{r setup, error=FALSE, message=FALSE, warning=FALSE, include=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)){install.packages("BiocManager")}
if (!require("GEOmetadb", quietly = TRUE)){install.packages("GEOmetadb")}
if (!requireNamespace("RSQLite", quietly = TRUE)){install.packages("RSQLite")}
if (!requireNamespace("readr", quietly = TRUE)){install.packages("readr")}
if (!requireNamespace("AnnotationDbi", quietly = TRUE)){install.packages("AnnotationDbi")}
if (!requireNamespace("org.Hs.eg.db", quietly = TRUE)){BiocManager::install("org.Hs.eg.db")}
if (!requireNamespace("wordcloud", quietly = TRUE)){install.packages("wordcloud")}
if (!requireNamespace("edgeR", quietly = TRUE)){install.packages("edgeR")}

```

```{r library, error=FALSE, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
library(BiocManager)
library(GEOmetadb)
library(RSQLite)
library(readr)
library(utils)
library(base)
library(readxl)
library(biomaRt)
library(dplyr)
library(knitr)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(wordcloud)
library(stringr)
library(edgeR)
})
```

## Results from previous assigment

### Extract data

Download the data.
```{r file names, error=FALSE, message=FALSE, warning=FALSE}
sfiles = getGEOSuppFiles('GSE153873')
fnames = rownames(sfiles)

new_fnames <- c()

for (string in fnames) {
  new_string <- sub(".*/Assignment2", "/Assignment2", string)
  new_fnames <- c(new_fnames, new_string)
}
```

```{r file glimpse, error=FALSE, message=FALSE, warning=FALSE}
b3 = read.delim(fnames[3], header = TRUE)
dim(b3)
```
### Gene Duplication Check
```{r duplication check, message=FALSE}
length(b3$refGene)-length(unique(b3$refGene)) 

# double checking there are no duplicated genes
summarized_gene_counts <- sort(table(b3$refGene),
decreasing = TRUE)
```

### Check and remove low count genes
```{r Remove low count, message=FALSE, warning=FALSE}

cpms = cpm(b3[, 2:dim(b3)[2]])
keep30 = rowSums(cpms >1) >=30
b3_filtered30 = b3[keep30,]
```

### Normalization - EdgeR Container module and normalization with DEGList
```{r EdgeR, message=FALSE, error=FALSE, warning=FALSE}
filtered_data_matrix <- as.matrix(b3_filtered30[, 2:dim(b3_filtered30)[2]])
rownames(filtered_data_matrix) <- b3_filtered30$refGene

exp_condition <- c("AD", "Old", "Young")
exp_condition_vec <- str_extract(colnames(b3_filtered30), str_c(exp_condition , collapse="|"))

d = DGEList(counts=filtered_data_matrix, group = exp_condition_vec[2:length(exp_condition_vec)])
d = calcNormFactors(d)
normalized_data <- cpm(d)

dim(d)
```
## Differential Gene Expression

## Thresholded Over Representation Analysis

## Citation

- Used code snippets from Lecture 4 and 5 of BCB420, written by Professor Ruth Isserlin.
- RMarkdown Cookbook: https://bookdown.org/yihui/rmarkdown-cookbook/figure-size.html

```{r reference, collapse=TRUE}
citation("BiocManager")
citation("GEOmetadb")
citation("RSQLite")
citation("readr")
citation("utils")
citation("base")
```

