---
title: "Assignment 1"
author: "Anna Lai"
date: "February 14, 2023"
output:
  html_document:
    toc: yes
    df_print: paged
---

# Exploration of gene set associated with the Alzheimer’s disease

## Setup

Package installation commands hided. 
```{r setup, error=FALSE, message=FALSE, warning=FALSE, include=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)){install.packages("BiocManager")}
if (!require("GEOmetadb", quietly = TRUE)){install.packages("GEOmetadb")}
if (!requireNamespace("RSQLite", quietly = TRUE)){install.packages("RSQLite")}
if (!requireNamespace("readr", quietly = TRUE)){install.packages("readr")}
if (!requireNamespace("AnnotationDbi", quietly = TRUE)){install.packages("AnnotationDbi")}
if (!requireNamespace("org.Hs.eg.db", quietly = TRUE)){BiocManager::install("org.Hs.eg.db")}
if (!requireNamespace("wordcloud", quietly = TRUE)){install.packages("wordcloud")}
if (!requireNamespace("edgeR", quietly = TRUE)){install.packages("edgeR")}

```

```{r library, error=FALSE, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
library(BiocManager)
library(GEOmetadb)
library(RSQLite)
library(readr)
library(utils)
library(base)
library(readxl)
library(biomaRt)
library(dplyr)
library(knitr)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(wordcloud)
library(stringr)
library(edgeR)
})
```

## Data cleaning


### Information about the dataset

As the data description in GEO, there are 30 samples. Basic information about the dataset is shown below.

```{r data info, error=FALSE, message=FALSE, warning=FALSE}
metadata <- getGEO("GSE153873")

dim(metadata$GSE153873_series_matrix.txt.gz)

metadata$GSE153873_series_matrix.txt.gz$contact_institute[1]
metadata$GSE153873_series_matrix.txt.gz$contact_city[1]
metadata$GSE153873_series_matrix.txt.gz$contact_country[1]

metadata$GSE153873_series_matrix.txt.gz$status[1]
metadata$GSE153873_series_matrix.txt.gz$submission_date[1]
metadata$GSE153873_series_matrix.txt.gz$last_update_date[1]

metadata$GSE153873_series_matrix.txt.gz$organism_ch1[1]
metadata$GSE153873_series_matrix.txt.gz$source_name_ch1[1]
metadata$GSE153873_series_matrix.txt.gz$`tissue:ch1`[1]
metadata$GSE153873_series_matrix.txt.gz$extract_protocol_ch1[1]
```

### Information about the platform
```{r platform, message=FALSE}
gse <- getGEO("GSE153873",GSEMatrix=FALSE)
current_gpl <- names(GPLList(gse))[1]
current_gpl_info <- Meta(getGEO(current_gpl))
current_gpl_info$`
title`
```

**Q1 What are the controls and the test conditions of the dataset?** There are 12 disease samples, 10 samples from old individuals, and 8 from young individuals. 

**Q2 Why is this data set of interest to you?** Alzheimer’s disease (AD) has long thought to be an age related disease with an onset age at around 65 years old. By comparing the brain tissue samples of young, adult, and diseased brain, it can yield convincing results on certain genes that regulates the disease. 

```{r sample info, message=FALSE}
exp_con <- data.frame(title = metadata$GSE153873_series_matrix.txt.gz$title,
           disease_state = metadata$GSE153873_series_matrix.txt.gz$`disease state:ch1`)
exp_con

count <- table(exp_con$disease_state)
count
```

### Extract data

```{r file names, error=FALSE, message=FALSE, warning=FALSE}
sfiles = getGEOSuppFiles('GSE153873')
fnames = rownames(sfiles)

new_fnames <- c()

for (string in fnames) {
  new_string <- sub(".*/Assignment1", "/Assignment1", string)
  new_fnames <- c(new_fnames, new_string)
}
new_fnames
```

There are 3 supplementary files from this paper. In file 2 and 2, 31 columns are observed in each table, the first column is the Gene identifier. 

```{r file glimpse, error=FALSE, message=FALSE, warning=FALSE}
b1 = read_excel(fnames[1])
b2 = read.delim(fnames[2], header = TRUE)
b3 = read.delim(fnames[3], header = TRUE)

head(b1)
head(b2)
head(b3)
dim(b1)
dim(b2)
dim(b3)
```
### Data and Sample condition

File 3 is the file that contains the expression data of the genes of interest. Proceed with file 3. There are 271135 genes in total in the data set.

```{r map col names, message=FALSE}
b3_cols<- colnames(b3[2:31])
b3_cols_format <- c()
for (string in b3_cols) {
  new_string <- gsub("X(\\d+).(\\d+)(\\w+).(\\w+)", "\\1-\\2\\3-\\4", string)
  b3_cols_format <- c(b3_cols_format, new_string)
}
exp_con_merged <- merge(exp_con, data.frame(b3_cols = b3_cols, title = b3_cols_format), by = "title")
exp_con_merged
```

### Gene Duplication

**Q Were there expression values that were not unique for specific genes?** It is likely that the expression values and the symbols have been cleaned before being deposited into the GEO. Hence, from tables before, there doesn't seem to have expression values that were not unique.

**Q How did you handle replicates in this data?** There are no replicates in this data set.
```{r duplication check, message=FALSE}
length(b3$refGene)-length(unique(b3$refGene)) 

# double checking there are no duplicated genes
summarized_gene_counts <- sort(table(b3$refGene),
decreasing = TRUE)
kable(table(b3$refGene)[1:5], format="html")
kable(summarized_gene_counts[which(summarized_gene_counts>1)[1:5]],format="html")
```

### HUGO symbol mapping
The data is already using HUGO symbol. Hence for the purpose of the assignment, I mapped them to ensembl_gene_id. From the code snippet below we conclude that there are 3592 genes without ensembl_gene_id, and 1466 genes maps to more than 1 ensembl_gene_id.

**Q Were there any expression values that could not be mapped to current HUGO symbols?** NO. The dataset uses HUGO notation. 

```{r HUGO mapping, message=FALSE}

ensembl <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")

genes_to_ensembl <- getBM(attributes = c("hgnc_symbol", "ensembl_gene_id"), 
                           filters = "hgnc_symbol", 
                           values = b3$refGene, 
                           mart = ensembl)

# number of genes that are not mapped to Ensembl ID
included_in_both <- intersect(b3$refGene, genes_to_ensembl$hgnc_symbol) 
length(b3$refGene)-length(included_in_both)

# genes that mapped to more than 1 ensembl_gene_id
dup_rows <- duplicated(genes_to_ensembl$hgnc_symbol) | duplicated(genes_to_ensembl$hgnc_symbol, fromLast = TRUE)
duplicated_df <- genes_to_ensembl[dup_rows, ]
length(unique(duplicated_df$hgnc_symbol))
```

### Check and remove low count genes
***Question 5, how many outliers are removed?*** From the code below, we removed 9713 low count genes. 
```{r Remove low count, message=FALSE, warning=FALSE}

cpms = cpm(b3[, 2:dim(b3)[2]])
keep10 = rowSums(cpms >1) >=10
b3_filtered10 = b3[keep10,]

keep30 = rowSums(cpms >1) >=30
b3_filtered30 = b3[keep30,]

dim(b3)[1] - dim(b3_filtered10)[1]
dim(b3)[1] - dim(b3_filtered30)[1]

dim(b3_filtered10)[1] - dim(b3_filtered30)[1]
```

## Distribution of Data Before normalization

## GO term mapping
Mapping to the GO terms to observe the overall genes included in the data set after filtering out the low count genes.
```{r GO mapping, message=FALSE, warning=FALSE, error=FALSE}
mapped_terms <- select(org.Hs.eg.db, keys=b3_filtered30$refGene, keytype="SYMBOL", columns= c("GENENAME", "ONTOLOGY" ,"PATH"))

freq_df <- data.frame(table(mapped_terms$PATH))

kegg_name <- rep("Not Found in KEGG", length(freq_df$Var1))

for (i in 1:length(freq_df$Var1)) {
  kegg_pathway <- paste("hsa", freq_df$Var1[i], sep = "")
  url <- paste("http://rest.kegg.jp/get/", kegg_pathway, sep = "")
  
    tryCatch(
    {
      lines <- readLines(url, warn = FALSE)
       match <- str_match(lines[2], "NAME\\s+(.*)\\s+-\\s+(.*)")
      if (!is.na(match[1,2])) {
        pathway_name <- match[1,2]
        organism_name <- match[1,3]
        kegg_name[i]<- pathway_name
      }
    },
    error = function(e) {
      message(paste0("ERROR: ", e$message))
    }
  )
  
}
# kegg_name
```

```{r GO mapping Word Cloud, message=FALSE, fig.height= 8, fig.width=8}
wordcloud(words = kegg_name, freq = freq_df$Freq, min.freq = 1, max.words = 350, random.order = FALSE, rot.per = 0.35,
          colors = brewer.pal(28, "Dark2"))
```

### Box plot
At this stage, there is no clear separation between the overall expression data between the three sample groups: young, old, AD diseased. There are a lot of outliers in the data as shown in the box plot.
```{r box plot, message=FALSE, error=FALSE, warning=FALSE}
b3plot <- log2(cpm(b3_filtered30[, 2:dim(b3_filtered30)[2]]))
boxplot(b3plot, xlab = "Sample", ylab = "log2 CPM",
las = 2, cex = 0.5, cex.lab = 0.5,
cex.axis = 0.5, main = "Box plot of RNA expression of 30 samples")

abline(h = median(apply(b3plot, 2, median)),
col = "blue3", lwd = 2, lty = "dashed")

```
### Density plot
```{r density, message=FALSE, error=FALSE, warning=FALSE, fig.height= 8, fig.width=10}
plot_Density <- function(b3plot){
counts_density <- apply(b3plot, 2, density)

#calculate the limits across all the samples
xlim <- 0; ylim <- 0
for (i in 1:length(counts_density)) {
xlim <- range(c(xlim, counts_density[[i]]$x));
ylim <- range(c(ylim, counts_density[[i]]$y))
}
cols <- rainbow(length(counts_density))
ltys <- rep(1, length(counts_density))

#plot the first density plot to initialize the plot
plot(counts_density[[1]], xlim=xlim, ylim=ylim, type="n",
ylab="Smoothing density of log2-CPM", main="Density plot of RNA expression data after low count removal", cex.lab = 0.8)

#plot each line
for (i in 1:length(counts_density)) lines(counts_density[[i]], col=cols[i], lty=ltys[i])

#create legend
legend("right", colnames(b3plot),
col=cols, lty=ltys, cex=0.75,
border ="blue", text.col = "black",
merge = TRUE, bg = "gray90")
}

plot_Density(b3plot)
```

## Normalization

### MA Plot before nomrmalization
As my data set is sampled based, I will normalize the data with Trimmed Mean of M-values (TMM)
 method. I categorized the data in to three conditions according to the publication and picked number 3 randomly to plot the MA plot. From the graphics, the selected 3 samples show that the log expression of gene differs between the samples with different conditions, especially between young and old brains. 
 
```{r MA Plot, message=FALSE, error=FALSE, warning=FALSE, fig.height= 8, fig.width=10}
sample_col <- colnames(b3_filtered30)
young_cols <- sample_col[grep("Young", sample_col)]
ad_cols <- sample_col[grep("AD", sample_col)]
old_cols <- sample_col[grep("Old", sample_col)]
other_cols <- sample_col[!(sample_col %in% c(young_cols, ad_cols, old_cols))]

par(mfrow = c(2, 2))
limma::plotMA(log2(b3_filtered30[, c(ad_cols[3], old_cols[3])]), ylab="M - ratio log expression", 
       main="AD Diseased sample 3 vs Old sample 3")
       abline(h = 0, col = "blue3", lwd = 2, lty = "dashed")
limma::plotMA(log2(b3_filtered30[, c(ad_cols[3], young_cols[3])]), ylab="M - ratio log expression", 
       main="AD Diseased sample 3 vs Young sample 3")
       abline(h = 0, col = "blue3", lwd = 2, lty = "dashed")
limma::plotMA(log2(b3_filtered30[, c(young_cols[3], old_cols[3])]), ylab="M - ratio log expression", 
       main="Young sample 2 vs Old sample 2")
       abline(h = 0, col = "blue3", lwd = 2, lty = "dashed")
```
### EdgeR Container module and normalization with DEGList
```{r EdgeR, message=FALSE, error=FALSE, warning=FALSE}
filtered_data_matrix <- as.matrix(b3_filtered30[, 2:dim(b3_filtered30)[2]])
rownames(filtered_data_matrix) <- b3_filtered30$refGene

exp_condition <- c("AD", "Old", "Young")
exp_condition_vec <- str_extract(colnames(b3_filtered30), str_c(exp_condition , collapse="|"))

d = DGEList(counts=filtered_data_matrix, group = exp_condition_vec[2:length(exp_condition_vec)])
d = calcNormFactors(d)
normalized_data <- cpm(d)

dim(d)
```
**Q What's the final coverage of your data?** The final coverage is 13986 genes. 

### Box plot after normalization
After normalization, the median of each sample differ less to each other than before, which makes the data more comparable. The number of outlier genes (above the maxium point in the boxplots) in each sample remains unchanged. Note that the box plots seemed to have normal distributions as the medians are located at the center of each box. 

```{r After Box plots, message=FALSE, error=FALSE, warning=FALSE, fig.height= 12, fig.width=8}
par(mfrow = c(2, 1))

boxplot(b3plot, xlab = "Sample", ylab = "log2 CPM",
las = 2, cex = 0.5, cex.lab = 0.5,
cex.axis = 0.5, main = "Box plot of RNA expression of 30 samples before normalization")

abline(h = median(apply(b3plot, 2, median)),
col = "blue3", lwd = 2, lty = "dashed")

boxplot(log2(normalized_data), xlab = "Sample", ylab = "log2 CPM",
las = 2, cex = 0.5, cex.lab = 0.5,
cex.axis = 0.5, main = "Box plot of RNA expression of 30 samples after normalization")

abline(h = median(apply(b3plot, 2, median)),
col = "cyan", lwd = 3, lty = "dashed")
```

### Density plots after normalization
There is no significant change in the density plot after normalization.
```{r n density plots, message=FALSE, error=FALSE, warning=FALSE, fig.height= 8, fig.width=10}
par(mfrow = c(1, 2))

plot_Density(b3plot)
plot_Density(log2(normalized_data))

```

### MDS Plot after normalization
```{r Group, message=FALSE, error=FALSE, warning=FALSE}
samples <- data.frame(lapply(colnames(normalized_data)[1:31], 
        FUN=function(x){unlist(strsplit(x, 
                        split = "\\."))}))
colnames(samples) <- colnames(normalized_data)[1:31]
rownames(samples) <- c("sample_number", "A_T", "disease_type")
samples <- data.frame(t(samples))
samples <- samples[1:30, ]
```

From the MDS Plot, three clusters are observed. The green one representing the AD population clusters in the center, while the old population clusters at the left hand side of the graph. Red, representing young population in the sample, scatter across x a-xis, but does not have a y value higher than 0.5.

```{r MDS plot, message=FALSE, error=FALSE, warning=FALSE}
plotMDS(d, labels=rownames(samples),
col = c("darkgreen","blue", "red")[factor(samples$disease_type)],
main="MDS plot of Young, Old, AD samples")
```
### BCV plot
Biological coefficient of variation(BCV) plot. Genes with more counts should have less variance than the high counts as discussed in class. However, it is not the case in this data set from the BCV generated below. A high number of outliers with high count might explain the behavior of this resulting plot. 
```{r}
# warning shown when model design is included in the calculation of dispersion. Omitted for now. 
model_design <- model.matrix(~samples$disease_type +   0)
bcv_d <- estimateDisp(d, model_design)

plotBCV(bcv_d, col.tagwise = "black",col.common = "red",
        main= "BCV plot for Old, Young, AD type cells")
```

### Mean vairance plot
Each dot on the graph represents one gene. 
```{r}
plotMeanVar(bcv_d, show.raw.vars = TRUE, show.tagwise.vars=TRUE,
show.ave.raw.vars = TRUE,
NBline=TRUE,
show.binned.common.disp.vars = TRUE,
main="Mean Variance plot of expression level of genes")
```

## References

### Ciatations

- Used code snippets from Lecture 4 and 5 of BCB420, written by Professor Ruth Isserlin.
- RMarkdown Cookbook: https://bookdown.org/yihui/rmarkdown-cookbook/figure-size.html

```{r reference, collapse=TRUE}
citation("BiocManager")
citation("GEOmetadb")
citation("RSQLite")
citation("readr")
citation("utils")
citation("base")
```

