---
title: "Assignment 1"
author: "Anna Lai"
date: "February 14, 2023"
output:
  html_document:
    toc: yes
    df_print: paged
---

# Exploration of gene set associated with the Alzheimer’s disease

## Setup

Package installation commands hided. 
```{r setup, error=FALSE, message=FALSE, warning=FALSE, include=FALSE}

# remove.packages("BiocManager")
if (!requireNamespace("BiocManager", quietly = TRUE))
install.packages("BiocManager")
if (!require("GEOmetadb", quietly = TRUE))
    install.packages("GEOmetadb")
if (!requireNamespace("RSQLite", quietly = TRUE))
install.packages("RSQLite")
if (!requireNamespace("readr", quietly = TRUE))
install.packages("readr")
```

```{r library, error=FALSE, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
library(BiocManager)
library(GEOmetadb)
library(RSQLite)
library(readr)
library(utils)
library(base)
library(readxl)
})
```

## Data cleaning


### Information about the dataset

As the data description in GEO, there are 30 samples. Basic information about the dataset is shown below.

```{r data info, error=FALSE, message=FALSE, warning=FALSE}
metadata <- getGEO("GSE153873")

dim(metadata$GSE153873_series_matrix.txt.gz)

metadata$GSE153873_series_matrix.txt.gz$contact_institute[1]
metadata$GSE153873_series_matrix.txt.gz$contact_city[1]
metadata$GSE153873_series_matrix.txt.gz$contact_country[1]

metadata$GSE153873_series_matrix.txt.gz$status[1]
metadata$GSE153873_series_matrix.txt.gz$submission_date[1]
metadata$GSE153873_series_matrix.txt.gz$last_update_date[1]

metadata$GSE153873_series_matrix.txt.gz$organism_ch1[1]
metadata$GSE153873_series_matrix.txt.gz$source_name_ch1[1]
metadata$GSE153873_series_matrix.txt.gz$`tissue:ch1`[1]
metadata$GSE153873_series_matrix.txt.gz$extract_protocol_ch1[1]
```

### Information about the platform
```{r platform, message=FALSE}
gse <- getGEO("GSE153873",GSEMatrix=FALSE)
current_gpl <- names(GPLList(gse))[1]
current_gpl_info <- Meta(getGEO(current_gpl))
current_gpl_info$`
title`
```

**Q1 What are the controls and the test conditions of the dataset?** There are 12 disease samples, 10 samples from old individuals, and 8 from young individuals. 

```{r sample info, message=FALSE}
exp_con <- data.frame(title = metadata$GSE153873_series_matrix.txt.gz$title,
           disease_state = metadata$GSE153873_series_matrix.txt.gz$`disease state:ch1`)
exp_con

count <- table(exp_con$disease_state)
count
```

### Extract data

```{r file names, error=FALSE, message=FALSE, warning=FALSE}
sfiles = getGEOSuppFiles('GSE153873')
fnames = rownames(sfiles)

new_fnames <- c()

for (string in fnames) {
  new_string <- sub(".*/Assignment1", "/Assignment1", string)
  new_fnames <- c(new_fnames, new_string)
}
new_fnames
```

There are 3 supplementary files from this paper. In file 2 and 2, 31 columns are observed in each table, the first column is the Gene identifier. 

```{r file glimpse, error=FALSE, message=FALSE, warning=FALSE}
b1 = read_excel(fnames[1])
b2 = read.delim(fnames[2], header = TRUE)
b3 = read.delim(fnames[3], header = TRUE)

head(b1)
head(b2)
head(b3)
dim(b1)
dim(b2)
dim(b3)
```
### Data and Sample condition

File 3 is the file that contains the expression data of the genes of interest. Proceed with file 3. There are 271135 genes in total in the data set.

```{r map col names, message=FALSE}
b3_cols<- colnames(b3[2:31])
b3_cols_format <- c()
for (string in b3_cols) {
  new_string <- gsub("X(\\d+).(\\d+)(\\w+).(\\w+)", "\\1-\\2\\3-\\4", string)
  b3_cols_format <- c(b3_cols_format, new_string)
}
exp_con <- merge(exp_con, data.frame(b3_cols = b3_cols, title = b3_cols_format), by = "title")
exp_con_merged
```

### Gene Duplication
```{r duplication check, message=FALSE}
length(b3$refGene)-length(unique(b3$refGene)) 

# double checking there are no duplicated genes
summarized_gene_counts <- sort(table(b3$refGene),
decreasing = TRUE)
kable(table(b3$refGene)[1:5], format="html")
kable(summarized_gene_counts[which(summarized_gene_counts>1)[1:5]],format="html")
```

### HUGO symbol mapping
```{r HUGO mapping, message=FALSE}

```

### Check and remove outliers
```{r}

```


## Normalization
```{r}

```


## References
```{r reference, collapse=TRUE}
citation("BiocManager")
citation("GEOmetadb")
citation("RSQLite")
citation("readr")
citation("utils")
citation("base")
```

